---
title: "Working with raster data"
description: |
  Visualizing marine species richness in California
author:
  - name: Grace Brofman
    url: {}
date: 03-07-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: Code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = FALSE)
```

### **Introduction:**
Aquamaps is a map-based tool which provides data and visualizations for the world's marine species. Here, cetacean data for a portion of California is used to visualize marine species richness. 

```{r, warning = FALSE, message = FALSE}
library(raster) # load this before tidyverse
library(tidyverse)
library(here)
library(sf)
library(fasterize)
library(rnaturalearth)
library(RColorBrewer)
```

```{r, warning = FALSE, message = FALSE}
# load and instpect the data
my_files <- list.files(path = "ca_cetaceans", full.names = TRUE)

tif_stack <- raster::stack(my_files)
tif_stack

plot(tif_stack)


```

```{r, warning = FALSE, message = FALSE}


# CA shapefile

state <- read_sf("ca_state_border",
                    layer = "CA_State_TIGER2016")

#check projection
st_crs(state) # WGS 84 / 3857

# projections differ, so transform one CRS to the other:
state <- st_transform(state, st_crs(tif_stack))

st_crs(state) # 9122
```


```{r, warning = FALSE, message = FALSE}
# setting threshold?

threshold_function <- function(x, thresh = .75) {
  y <- ifelse(x >= thresh, 1, NA)
  return(y)
}

species_thresh <- calc(tif_stack, fun = threshold_function)

# adding species richness in each raster

species_richness <- raster::calc(species_thresh, fun = sum, na.rm = TRUE)
plot(species_richness)

# mask species_richness with coastline data -> also unnecessary?
#mask(species_richness, coastline110) %>% 
#  plot()

# then convert to a dataframe

richness_df <- raster::rasterToPoints(species_richness) %>% 
  as.data.frame()

richness_df[richness_df == 0] <- NA # replacing zero values with NA so that it doesn't show up when plotted
```

```{r, warning = FALSE, include = TRUE, fig.cap = "**Figure 1.** Map of species richness of Cetaceans along the California coast. Richness is a measure of aggregate species presence as opposed to a single species. A threshold of 0.75 was chosen in order to observe greater detail in species richness patterns, while maintaining coverage for areas of lesser species richness ."}
# then plot

ggplot() +
  geom_raster(data = richness_df,
              aes(x = x, y = y, fill = layer)) +
  geom_sf(data = state, fill = "snow2") +
  coord_sf(xlim = c(-126, -114), ylim = c(31, 42)) + # defining coordinates
  theme_minimal() +
  labs(x = "longitude", y = "latitude",
       title = "Species richness of Cetaceans along California's coast",
       fill = "Species richness") +
  scale_fill_viridis_c(na.value = "white") 

```




#### **References**

**Kaschner, K., K. Kesner-Reyes, C. Garilao, J. Segschneider, J. Rius-Barile, T. Rees, and R. Froese.** 2019. AquaMaps: Predicted range maps for aquatic species. World wide web electronic publication, www.aquamaps.org, Version 10/2019.






