---
title: "Text wrangling and analysis"
description: |
  Word clouds and sentiment analysis for a silly youth magazine.
author:
  - name: Grace Brofman
    url: {}
date: 03-07-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: Code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, include = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidytext)
library(textdata)
library(pdftools)
library(ggwordcloud)
library(patchwork)
```
### **Introduction:**
For this task I used an unsavory and perhaps sexist magazine I found on the [(Internet Archive)](https://archive.org/details/texts). I chose to perform a sentiment analysis on the 2009 edition of **[(Just for Girls and Guys Youth Magazine)](https://archive.org/details/j4g09/page/n15/mode/2up)** by the Human Life Alliance. The first half (pages 1-8) is "just for girls", while the second half (pages 9-16) is "just for boys". The two halves are meant to give each side insight into the other gender. I wanted to see how the sentiments differed (if at all) between the "girls'" side and the "boys'" side.
```{r, include = FALSE, warning = FALSE, message = FALSE}
# upload text 
mag_text <- pdf_text("Just for Girls and Guys.pdf")

# tidy text
mag_tidy <- data.frame(mag_text) %>% 
  mutate(text_full = str_split(mag_text, pattern = "page_break")) %>%   # breaks at "page_break"
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full)) # trims excess whitespace

mag_df <- mag_tidy %>% 
  mutate(page = case_when(
    str_detect(text_full, pattern = "Beginning page") ~ text_full,
    TRUE ~ NA_character_
  )) %>% 
  fill(page) %>% 
  separate(col = page, into = c("pg", "no"), sep = "page ") %>% 
  mutate(page = as.numeric(as.character(no))) # changing page number from character to numberic in new column

```

```{r, include = FALSE, warning = FALSE, message = FALSE}
# convert to tokens using unnest_tokens()

mag_tokens <- mag_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-mag_text) %>% 
  filter(!word == "100") %>% 
  filter(!word == "10") %>% 
  filter(!word == "9") %>% 
  filter(!word == "8") %>% 
  filter(!word == "7") %>% 
  filter(!word == "6") %>% 
  filter(!word == "5") %>% 
  filter(!word == "4") %>% 
  filter(!word == "3") %>% 
  filter(!word == "2") %>% 
  filter(!word == "1")  %>% 
  filter(!word == "16") %>% 
  filter(!word == "2010") %>% 
  filter(!word == "2011") %>%
  filter(!word == "25")
  

# word count by page:
mag_wordcount <- mag_tokens %>% 
  count(page, word)

# remove stop words from mag_tokens
mag_nonstop_words <- mag_tokens %>% 
  anti_join(stop_words)

nonstop_counts <- mag_nonstop_words %>% 
  count(page, word)

nonstop_counts

# top 5 words by page
top_5_words <- nonstop_counts %>% 
  group_by(page) %>% 
  arrange(-n) %>% 
  slice(1:5)

ggplot(data = top_5_words,
       aes(x = word, y = n)) +
  geom_col(fill = "blue") +
  facet_wrap(~page, scales = "free") +
  coord_flip()


```

### **Part 1:** Word count visualization

```{r, warning = FALSE, message = FALSE}
# girl-focused pages are 1-8. boy-focused pages are 9-16.
girl_top_50 <- nonstop_counts %>% 
  filter(page == c(1:8)) %>% 
  arrange(-n) %>% 
  slice(1:50)

boy_top_50 <- nonstop_counts %>% 
  filter(page == c(9:16)) %>% 
  arrange(-n) %>% 
  slice(1:50)

girl_cloud <- ggplot(data = girl_top_50,
                     aes(label = word)) +
  geom_text_wordcloud(aes(color = n, size = n)) +
  scale_size_area(max_size = 8) +
  labs(title = "50 most common words in Just for Girls") 


boy_cloud <- ggplot(data = boy_top_50,
                     aes(label = word)) +
  geom_text_wordcloud(aes(color = n, size = n)) +
  scale_size_area(max_size = 8) +
  labs(title = "50 most common words in Just for Boys") 




```

```{r}
girl_cloud

```

**Figure 1:** Word cloud for the 50 most common words in "Just for Girls", pages 1-8. Word size is representative of frequency.

```{r}
boy_cloud
```

**Figure 2:** Word cloud for the 50 most common words in "Just for Boys", pages 9-16. Word size is representative of frequency.

### **Part 2:** Sentiment analysis
```{r, warning = FALSE, message = FALSE}
# sentiment analysis using AFINN lexicon 
mag_afinn <- mag_nonstop_words %>%  # finding common words beteen mag_nonstop_words and afinn lexicon
  inner_join(get_sentiments("afinn"))

# for each page, how many times does each sentiment score show up?
afinn_counts <- mag_afinn %>% 
  count(page, value)

afinn_means <- mag_afinn %>% 
  group_by(page) %>% 
  summarize(mean_afinn = mean(value))

plot_2 <- ggplot(data = afinn_means,
       aes(x = page,
           y = mean_afinn)) +
  geom_col(aes(fill = mean_afinn)) +
  scale_fill_gradient(low = "deeppink",
  high = "dodgerblue2") +
  geom_vline(xintercept = 8, linetype = "dashed", color = "darkgray") +
  labs(x = "Page number", y = "Score", title = "Mean AFINN score per page", fill = "Score range") +
  coord_flip() +
  theme_minimal()

plot_2

# girl-targeted pages are 1-8, boy-targeted pages are 9-16. it seems that in general, the boy-targeted pages are associated with a higher positivity than the girl-targeted pages
```

**Figure 3:** Sentiment analysis of Just for Girls and Boys Youth Magazine using the AFINN lexicon. Pages 1-8 are "Just for Girls" while pages 9-16 are "Just for Boys". 


### **References**
**Human Life Alliance**. "Just for Girl and Guys Youth Magazine." 2009: https://archive.org/details/j4g09/mode/2up
